---
title: "Combined Methods Approach in deconvMe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combined Methods Approach in deconvMe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Combined Methods Approach in deconvMe

The `deconvolute_combined()` function in deconvMe allows you to run multiple cell-type deconvolution methods simultaneously and create aggregated results. This approach can help reduce method-specific biases and provide more robust estimates of cell-type proportions.

## Basic Usage

```{r, message=FALSE, warning=FALSE}
library(deconvMe)
library(minfiData)
options(matrixStats.useNames.NA = "deprecated")

# Example data
methyl_set <- minfiData::MsetEx

# Run multiple methods
result_combined <- deconvolute_combined(
  methyl_set = methyl_set,
  methods = c('epidish', 'houseman', 'methylcc'),
  array = '450k'
)
```

## Understanding the Output

The combined results include both individual method results and aggregated estimates:

```{r}
# View the structure
head(result_combined)

# Check available methods
unique(result_combined$method)

# Check available cell types
unique(result_combined$celltype)
```

## Cell-Type Standardization

The function automatically standardizes cell-type names across methods:

```{r}
# Example of cell-type mapping
library(dplyr)

# Show how cell types are mapped
cell_type_mapping <- data.frame(
  Original = c("CD8T", "CD4T", "B", "NK", "Mono", "Neu", "Unknown"),
  Standardized = c("T cell CD8+", "T cell CD4+", "B cell", "NK cell", "Monocyte", "Neutrophil", "other")
)
knitr::kable(cell_type_mapping, caption = "Cell-type standardization mapping")
```

## The "Other" Category

The "other" category includes:
- Cell types not in the standardized mapping
- Method-specific cell types without equivalents
- Rare or specialized cell populations

This category should be interpreted carefully, as it may represent:
- True rare cell types
- Method-specific artifacts
- Incomplete cell-type coverage

## Visualizing Combined Results

```{r, fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
# Create visualization of individual vs aggregated results
library(ggplot2)

# Filter for one sample to show the comparison
sample_data <- result_combined[result_combined$sample == result_combined$sample[1], ]

ggplot(sample_data, aes(x = celltype, y = value, fill = method)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Individual Methods vs Aggregated Results",
       subtitle = paste("Sample:", sample_data$sample[1]),
       y = "Estimated Fraction",
       x = "Cell Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Limitations and Considerations

### 1. Method Heterogeneity
Different methods use different algorithms and reference datasets, which may not be directly comparable. Consider:
- Whether methods are validated for your tissue type
- The quality and relevance of reference datasets
- Algorithm-specific assumptions

### 2. Equal Weighting
The current implementation gives equal weight to all methods. This may not be optimal if:
- Some methods are more reliable for your specific use case
- Methods have different levels of validation
- You have prior knowledge about method performance

### 3. Missing Data
If a method fails to estimate a particular cell type, it may be excluded from aggregation, potentially biasing results.

### 4. Cell-Type Coverage
Methods may detect different sets of cell types, leading to incomplete coverage in aggregated results.

## Quality Control
```{r}
# Check for consistency across methods
consistency_check <- result_combined %>%
  group_by(sample, celltype) %>%
  summarise(
    mean_value = mean(value),
    sd_value = sd(value),
    cv = sd_value / mean_value,  # coefficient of variation
    .groups = 'drop'
  ) %>%
  filter(!is.na(mean_value))

# Identify cell types with high variability across methods
high_variability <- consistency_check %>%
  filter(cv > 0.5)  # arbitrary threshold
```

