---
title: "Getting Started with deconvMe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with deconvMe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting Started with deconvMe

Welcome to **deconvMe**! This vignette will help you get up and running with cell-type deconvolution for DNA methylation data using the package's unified interface and included methods.

## Installation

You can install deconvMe from GitHub using the [pak](https://github.com/r-lib/pak) package manager:

```{r, eval=FALSE}
# install the `pak` package manager
install.packages("pak")

pak::pkg_install("omnideconv/deconvMe")
```

## Example Usage

deconvMe can be applied directly to a methylSet from the minfi package, or you can apply each method separately on a beta matrix with Illumina CpG IDs.

Below, we demonstrate how to use the EpiDISH method with example data from minfi:

```{r, message=FALSE, warning=FALSE}
library(deconvMe)
library(minfi)
library(minfiData)

# use example data from Minfi
methyl_set <- minfiData::MsetEx
ratio_set <- minfi::ratioConvert(methyl_set)
beta_matrix <- minfi::getBeta(ratio_set)

# run EpiDISH for deconvolution of example data
result <- deconvMe::deconvolute(methyl_set = methyl_set, method = 'epidish')
```

## Viewing the Results

The result of the deconvolution is a table with the estimated cell-type fractions for each sample. You can view it directly as a nicely formatted table:

```{r}
knitr::kable(head(result), caption = "Estimated cell-type fractions for the first few samples")
```

For more details, see the package documentation and other vignettes! 

# Overview of Included Methods

Below is a template table summarizing the deconvolution methods included in deconvMe. This table lists each method, the type of algorithm it uses, the intended tissue types, and the type of data used to build the internal deconvolution reference.

| Name           | Algorithm Type(s)         | Intended Tissue Type(s)      | Reference Source Data                                 |
|:--------------|:-------------------------|:----------------------------|:------------------------------------------------------|
| EpiDISH        | Robust Partial Correlations (RPC), Constrained Projection (CP), Support Vector Regression (SVR) | Blood, Epithelial, Breast   | purified blood, epithelial, and non-epithelial data, extended with DNAse Hypersensitivity Sites from Roadmap and ENCODE |
| Houseman       | Constrained projection/Quadratic Programming (CP/QP) | Blood| Purified blood cell methylation profiles (IDOL-optimized) |
| MethylCC       | Constrained Linear model based on DMRs | Blood (in deconvMe, can be extended to other tissues)         | purified blood               |
| MethylResolver | Least Trimmed Squares (LTS) regression | Blood, cancer  | Purified blood methylation profiles (IDOL-optimized), extended with additional leukocytes |
| MethAtlas      | non-negative least squares (NNLS) | Blood, immune, tissue-wide | Comprehensive atlas of purified tissue and immune cell methylomes |

*This table can be expanded as new methods are added to the package.* 