---
title: "Getting Started with methyldeconv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with methyldeconv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting Started with methyldeconv

Welcome to **methyldeconv**! This vignette will help you get up and running with cell-type deconvolution for DNA methylation data using the package's unified interface and included methods.

## Installation

You can install methyldeconv from GitHub using the [pak](https://github.com/r-lib/pak) package manager:

```{r, eval=FALSE}
# install the `pak` package manager
install.packages("pak")

pak::pkg_install("omnideconv/methyldeconv")
```

## Example Usage

methyldeconv can be applied directly to a methylSet from the minfi package, or you can apply each method separately on a beta matrix with Illumina CpG IDs.

Below, we demonstrate how to use the EpiDISH method with example data from minfi:

```{r}
library(methyldeconv)
library(minfi)
library(minfiData)

# use example data from Minfi
methyl_set <- minfiData::MsetEx
ratio_set <- minfi::ratioConvert(methyl_set)
beta_matrix <- minfi::getBeta(ratio_set)

# run EpiDISH for deconvolution of example data
result <- methyldeconv::deconvolute(methyl_set = methyl_set, method = 'epidish')
```

## Viewing the Results

The result of the deconvolution is a table with the estimated cell-type fractions for each sample. You can view it directly as a nicely formatted table:

```{r}
knitr::kable(head(result), caption = "Estimated cell-type fractions for the first few samples")
```

For more details, see the package documentation and other vignettes! 