---
title: "Signature Matrices and Cross-Signature Deconvolution in deconvMe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Signature Matrices and Cross-Signature Deconvolution in deconvMe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Signature Matrices in deconvMe

deconvMe implements reference-based deconvolution methods that rely on *signature matrices* (reference matrices) to estimate cell-type proportions. A signature matrix is a table where each row corresponds to a CpG site and each column corresponds to a cell type, with entries representing expected methylation values for each CpG in each pure cell type.

## Accessing Signature Matrices

```{r, message=FALSE, warning=FALSE}
library(deconvMe)
library(dplyr)
library(tidyr)
library(ggplot2)
options(matrixStats.useNames.NA = "deprecated")

# Load example data
library(minfiData)
methyl_set <- minfiData::MsetEx
beta_matrix <- minfi::getBeta(minfi::ratioConvert(methyl_set))
```

All signature matrices follow a unified design - a `data.frame` with a `"CpGs"` column and cell types as additional columns:

```{r}
# Get signature matrices
epidish_sig <- get_epidish_signature_matrix("blood")
houseman_sig <- get_houseman_signature_matrix()
methatlas_sig <- get_methatlas_signature_matrix()
methylresolver_sig <- get_methylresolver_signature_matrix()

# Check structure
cat("EpiDISH:", dim(epidish_sig), "\n")
cat("Houseman:", dim(houseman_sig), "\n")
cat("MethAtlas:", dim(methatlas_sig), "\n")
cat("MethylResolver:", dim(methylresolver_sig), "\n")
```

## Custom CpG Subsets

You can use custom CpG subsets for deconvolution:

```{r, message=FALSE, warning=FALSE}
# Get overlapping CpGs between methods
custom_cpgs <- intersect(epidish_sig$CpGs, houseman_sig$CpGs)
cat("Overlapping CpGs:", length(custom_cpgs), "\n")

# Run with custom CpGs
result_custom <- run_epidish(beta_matrix, cpg_subset = custom_cpgs)
```

## Cross-Signature Deconvolution

deconvMe supports **cross-signature deconvolution**, allowing you to use signature matrices from one method with the deconvolution algorithm of another method.

### Supported Methods

| Method | Cross-Signature Support |
|--------|------------------------|
| EpiDISH | ✅ All methods |
| MethAtlas | ✅ All methods |
| MethylResolver | ✅ All methods |
| Houseman | ❌ Not supported |
| MethylCC | ❌ Not supported (no classical signature matrix) |

### Examples

```{r, message=FALSE, warning=FALSE}
# EpiDISH with Houseman signature
result_epidish_houseman <- run_epidish(beta_matrix, reference = houseman_sig)

# MethAtlas with EpiDISH signature
result_methatlas_epidish <- run_methatlas(beta_matrix, reference = epidish_sig)

# MethylResolver with MethAtlas signature
result_mr_methatlas <- run_methylresolver(
  beta_matrix, 
  reference = methatlas_sig,
  alpha = 1
)
```

### Comparing Results

```{r, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
# Compare EpiDISH with different signatures
result_epidish_blood <- run_epidish(beta_matrix, reference = "blood")$estF
result_epidish_houseman <- run_epidish(beta_matrix, reference = houseman_sig)$estF

# Rename cell types to unified names
colnames(result_epidish_blood) <- deconvMe::rename_cell_types(colnames(result_epidish_blood))
colnames(result_epidish_houseman) <- deconvMe::rename_cell_types(colnames(result_epidish_houseman))

result_epidish_blood_long <- result_epidish_blood |> 
  as.data.frame() |> 
  tibble::rownames_to_column('sample') |>
  tidyr::pivot_longer(-sample, names_to = 'celltype') |>
  dplyr::mutate('method'='EpiDISH')

result_epidish_houseman_long <- result_epidish_houseman |> 
  as.data.frame() |> 
  tibble::rownames_to_column('sample') |>
  tidyr::pivot_longer(-sample, names_to = 'celltype') |>
  dplyr::mutate('method'='Houseman')

results <- dplyr::bind_rows(result_epidish_blood_long, result_epidish_houseman_long) 

deconvMe::results_aggregated_boxplot(results) + 
  ggplot2::labs(fill = "signature")+
  ggtitle('Deconvolution with EpiDISH, with two different signatures')

```



