---
title: "Visualization Options in deconvMe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization Options in deconvMe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Visualization Options in deconvMe

deconvMe provides several functions to visualize the results of cell-type deconvolution analyses. This vignette demonstrates how to use these visualization tools with example data.

## Example Data

We'll use example data from the `minfi` and `minfiData` packages:

```{r, message=FALSE, warning=FALSE}
library(deconvMe)
library(minfi)
library(minfiData)

# Example data
methyl_set <- minfiData::MsetEx
ratio_set <- minfi::ratioConvert(methyl_set)
beta_matrix <- minfi::getBeta(ratio_set)

# Run deconvolution
result <- deconvMe::deconvolute(methyl_set = methyl_set, method = 'epidish')
result_multiple <- deconvMe::deconvolute_combined(methyl_set = methyl_set,
                                                      methods = c('epidish','methylcc'),
                                                      array = '450k')
```

## Barplot of Deconvolution Results

The `results_barplot()` function creates a barplot for each sample, showing the estimated cell-type fractions:

```{r, fig.height=4, fig.width=8}
deconvMe::results_barplot(result)
```

## Boxplot of Deconvolution Results

The `results_boxplot()` function creates a boxplot for each cell type, summarizing the distribution of estimated fractions across samples:

```{r, fig.height=4, fig.width=8}
deconvMe::results_boxplot(result)
```

## Aggregated Boxplot for Multiple Methods

If you run multiple methods using `deconvolute_combined()`, you can visualize the aggregated results with `results_aggregated_boxplot()`:

```{r, fig.height=4, fig.width=8}
deconvMe::results_aggregated_boxplot(result_multiple)
```

---

These visualization functions help you interpret and compare the results of different deconvolution methods. For more customization options, see the function documentation or explore the source code in the `R/visualization.R` file. 